\documentclass[main.tex]{subfiles}


\begin{document}

\chapter{Conclusion}

In this thesis, we discuss aspects of the decision process, in
particular, the mathematical formulation of optimisation problems and
algorithms used by numerical solvers to approximate solutions to such
problems.  When mathematicians and practitioners use models with
randomness, they often marginalise over random variables and then
mostly ignore the full distribution of the outcomes. The argument is
often that the marginalised value is a good representation of what the
decision maker cares, or should care, about.  It is the intention of
this thesis to highlight that the distribution of the outcomes for
relevant decisions can both guide the choice of optimisation
formulation and provide information regarding the impact algorithm
decisions have on modelled risk-attitudes.  The examples we provide to
motivate the work are mainly toy-problems related to pricing of retail
products, however, the considerations apply equally to other decision
processes with random outcomes.

\Cref{ch:onestage} illustrates that when the distributions are
uni-modal, or approximately normal, then formulating the optimisation
problem using mean and standard deviation can yield the same decision
as using super-quantiles. This is advantageous from both an
implementational and a computational point of view.
\Cref{ch:discrete_control,ch:cts_control} provide further examples of
how simplified formulations that ignore the randomness result in
marginalised outcomes that are practically equal to the values from
decisions based on the original formulation. A further investigation
of the distribution of outcomes shows that the simplified formulations
result in the best outcomes for half of the realisations of the
underlying random variables. In the remaining realisations, they
result in much worse outcomes, which we interpret as an algorithmic
decision that has impacted the modelled risk-attitude.  The
optimisation problems that we formulate in this thesis need to be
solved numerically using some optimisation routine using gradient
information. In \Cref{ch:objaccel} we propose a new algorithm that can
be combined with existing algorithms and we show, with several
numerical experiments, that it reduces the computational cost of
finding a solution with a prescribed accuracy.

The examples that we use to illustrate the impact of decisions and
distributions of outcomes are simplified problems of retail
applications. They provide a template for types of investigations that
can help the development of decision making infrastructure in
real-world applications.  In our examples, these investigations
conclude that simplified formulations reduce computational cost with
little impact on the objective, however, more complicated systems may
behave differently.  A natural avenue for further research is to
consider a real-world system where the demand models and the
distributional information is provided from a retailer. Within this
system one could then:
\begin{enumerate}
\item Investigate the typical distributions of objectives that arise
  based on this data;
\item propose two to four optimisation formulations; and
\item compare the computational cost, decisions, and distribution of
  outcomes that arise from the different formulations.
\end{enumerate}

The presence of competitors or external decisions that we do not
directly control is not discussed in this thesis. Competition is
believed to have a significant impact on the objectives by decision
maker's in business, and work to better understand such game theoretic
matters are important. A difficulty with theoretical investigations of
such situations is to that the researchers have too much control over
the problem: they decide apriori both how the underlying system
behaves and what policies the competitors can choose from.  We have
contributed to an experiment that runs a pricing and modelling
competition between several competitors \citep{geer2018dynamic}.  Each
competitor writes an algorithm that, only based on data, dynamically
prices a product over \num{1000} periods to maximise revenue without
knowing apriori what algorithms the competitors use. We believe this
provides a good way to test how the algorithms for decision-making
behaves in unexpected environments, and recommend that similar
competitions are run within the retailer before a new decision-making
system goes ``live''.



\biblio{} % Bibliography when standalone
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% TeX-command-extra-options: "--shell-escape"
%%% End:
