\documentclass[main.tex]{subfiles}

\begin{document}
Each of
\Cref{ch:onestage,ch:discrete_control,ch:cts_control,ch:objaccel}
contains original work.  The comparison of decisions from different
decision-making models in \Cref{ch:onestage} has not been conducted in
the revenue management community, and the multiobjective optimisation
package is written by the author.  The dynamic pricing problems in
\Cref{ch:discrete_control,ch:cts_control} use a different combination
of demand models, constraints, and profit functions from the existing
literature.  The investigation of the impact the optimal and
suboptimal pricing policies has on the profit distribution in these
chapters is more extensive than the marginalised comparisons that are
usually conducted. The solution and the asymptotic analysis of the
Hamilton--Jacobi--Bellman (HJB)
equations in \Cref{ch:cts_control} are original.  The optimisation
algorithm in \Cref{ch:objaccel} is an improvement of an existing idea,
but the theory and numerical tests are all original.

Much of the work and intuition behind this thesis is a result of
journal articles and software that the author has contributed to
throughout the D.Phil.\ research period.
The resulting articles and their relevance to the thesis are as follows.
\begin{itemize}
\item \citet{riseth2018erratum}:
  \textit{\usebibentry{riseth2018erratum}{title}};
  \begin{itemize}
    \begin{samepage}
    \item accepted in the \textit{\usebibentry{riseth2018erratum}{journal}};
    \item part of the work in \Cref{ch:onestage}.
    \end{samepage}
  \end{itemize}
\item \citet{riseth2017comparison}:
  \textit{\usebibentry{riseth2017comparison}{title}};
  \begin{itemize}
    \begin{samepage}
    \item submitted to the
      \textit{\usebibentry{riseth2017comparison}{submitted}};
    \item covers the majority of \Cref{ch:discrete_control}.
    \end{samepage}
  \end{itemize}
\item \citet{riseth2017dynamic}:
  \textit{\usebibentry{riseth2017dynamic}{title}};
  \begin{itemize}
    \begin{samepage}
    \item submitted to the
      \textit{\usebibentry{riseth2017dynamic}{submitted}};
    \item covers the majority of \Cref{ch:cts_control}.
    \end{samepage}
  \end{itemize}
\item \citet{riseth2017objective}:
  \textit{\usebibentry{riseth2017objective}{title}};
  \begin{itemize}
    \begin{samepage}
    \item submitted to
      \textit{\usebibentry{riseth2017objective}{submitted}};
    \item covers the majority of \Cref{ch:objaccel}.
    \end{samepage}
  \end{itemize}
\item \citet{mogensen2018optim}:
  \textit{\usebibentry{mogensen2018optim}{title}};
  \begin{itemize}
    \begin{samepage}
    \item published in the
      \textit{\usebibentry{mogensen2018optim}{journal}};
    \item software used in all chapters.
    \end{samepage}
  \end{itemize}
\item \citet{riseth2017operator}:
  \textit{\usebibentry{riseth2017operator}{title}};
  \begin{itemize}
    \begin{samepage}
    \item submitted to the
      \textit{\usebibentry{riseth2017operator}{submitted}};
    \item work on the data assimilation step of the decision process.
    \end{samepage}
  \end{itemize}
\item \citet{geer2018dynamic}:
  \textit{\usebibentry{geer2018dynamic}{title}};
  \begin{itemize}
    \begin{samepage}
    \item accepted in the
      \textit{\usebibentry{geer2018dynamic}{submitted}};
    \item initial work to guide further research.
    \end{samepage}
  \end{itemize}
\end{itemize}

\biblio{}
\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
